Score-based tests of measurement invariance: Use in practice

### 2014-01-25 frontiers/frontiers.Rnw
run a small simulation to make sure the frontiers.Rnw is set up correct.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@305 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

To conduct the analysis, we first load the data and keep only complete cases for simplicity 
(though the tests can be applied to incomplete data).
%
<<data>>=
data("YouthGratitude", package = "psychotools")
compcases <- apply(YouthGratitude[, 4:28], 1, function(x) all(x %in% 1:9))
yg <- YouthGratitude[compcases, ]
@
%
Next, we fit two models in \pkg{lavaan}: a one-factor model 
where loadings are restricted to be equal across age groups, 
and a one-factor model where loadings are free across age groups.
This allows us to test a hypothesis of partial measurement invariance
that was of interest to the original researchers (though, for ordinal
variables, all types of
measurement invariance can be examined via the tests described
previously).
By default, the code below sets the scale by fixing the first loading
to 1.
%
<<models>>=
restr <- cfa("f1 =~ gq6_1 + gq6_2 + gq6_3 + gq6_4 + gq6_5",
  data = yg, group = "agegroup", meanstructure = TRUE,
  group.equal = "intercepts")

@@

From these figures, one generally observes that the tests isolate 
the parameter violating measurement invariance.  Additionally, the tests 
have somewhat higher power to detect measurement invariance violations 
in the factor loading and factor covariance parameters, as opposed to the error variance
parameter\seeme{TODO:intercept discription}.  
Finally, simultaneous tests of all factor loadings, all intercepts or 
all error parameters result in decreased power, as compared to the situation
where one tests only the violating parameter.  This occurs because, in
testing a subset of parameters (only one of which violates measurement
invariance), we are effectively dampening the signal of a measurement
invariance violation. This ``dampening'' effect is more apparent for
the $\max \mathit{LM}_o$ statistic, because it involves a sum across
all tested parameters (see Equation~\eqref{eq:maxlmo}).
Conversely, $\mathit{WDM}_o$ takes the maximum over parameters
(Equation~\eqref{eq:wdm}), so that invariant parameters have no
impact on this statistic.

@@

\subsection{Results}
Results of
primary interest are conditions where the unmodeled loading violates
measurement invariance.  A subset of results is displayed in Figure~\ref{fig:sim21res}.  One can generally observe
that tests of the first loading and unique variance exhibited high
``power,'' which is actually a high Type I error rate here.  This Type
I error also demonstrated when testing all loadings and all unique
variances, as shown in the appendix.  Tests
associated with the factor 
covariance did not demonstrate this error, however\seeme(TODO:intercepts descrip).  In
terms of specific 
statistic performance, $\max \mathit{LM}_o$ and $\mathit{WDM}_o$ 
demonstrated higher Type I error than $\mathit{LM}_{uo}$ in each panel,
especially with increasing levels.  This is likely because the unmodeled
loading's non-invariance was monotonic with $V$; if it were not monotonic, we
would expect $\mathit{LM}_{uo}$ to have higher Type I error.

>>>>

To conduct the analysis, we first load the data and keep only complete cases for simplicity 
(though the tests can be applied to incomplete data).
%
<<data>>=
data("YouthGratitude", package = "psychotools")
compcases <- apply(YouthGratitude[, 4:28], 1, function(x) all(x %in% 1:9))
yg <- YouthGratitude[compcases, ]
@
%
Next, we fit two models in \pkg{lavaan}: a one-factor model 
where loadings are restricted to be equal across age groups, 
and a one-factor model where loadings are free across age groups.
This allows us to test a hypothesis of partial measurement invariance
that was of interest to the original researchers (though, for ordinal
variables, all types of
measurement invariance can be examined via the tests described
previously).
By default, the code below sets the scale by fixing the first loading
to 1.
%
<<models>>=
restr <- cfa("f1 =~ gq6_1 + gq6_2 + gq6_3 + gq6_4 + gq6_5",
  data = yg, group = "agegroup", meanstructure = TRUE,
  group.equal = "loadings")

@@

From these figures, one generally observes that the tests isolate 
the parameter violating measurement invariance.  Additionally, the tests 
have somewhat higher power to detect measurement invariance violations 
in the factor loading, factor covariance and intercept parameters, 
as opposed to the error variance parameter.  
Finally, simultaneous tests of all factor loadings, all intercepts or 
all error parameters result in decreased power, as compared to the situation
where one tests only the violating parameter.  This occurs because, in
testing a subset of parameters (only one of which violates measurement
invariance), we are effectively dampening the signal of a measurement
invariance violation. This ``dampening'' effect is more apparent for
the $\max \mathit{LM}_o$ statistic, because it involves a sum across
all tested parameters (see Equation~\eqref{eq:maxlmo}).
Conversely, $\mathit{WDM}_o$ takes the maximum over parameters
(Equation~\eqref{eq:wdm}), so that invariant parameters have no
impact on this statistic.

@@

\subsection{Results}
Results of
primary interest are conditions where the unmodeled loading violates
measurement invariance.  A subset of results is displayed in Figure~\ref{fig:sim21res}.  One can generally observe
that tests of the first loading and unique variance exhibited high
``power,'' which is actually a high Type I error rate here.  This Type
I error also demonstrated when testing all loadings and all unique
variances, as shown in the appendix.  Tests
associated with the factor 
covariance and intercept did not demonstrate this error, however.  In
terms of specific 
statistic performance, $\max \mathit{LM}_o$ and $\mathit{WDM}_o$ 
demonstrated higher Type I error than $\mathit{LM}_{uo}$ in each panel,
especially with increasing levels.  This is likely because the unmodeled
loading's non-invariance was monotonic with $V$; if it were not monotonic, we
would expect $\mathit{LM}_{uo}$ to have higher Type I error.
